Benchmarking neuromorphic systems with Nengo

### 2015-06-28 paper/paper.tex
First draft of paper done!
Needs work though.

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Bekolay {et~al.}}
\def\Authors{Trevor Bekolay\,$^{1,*}$,
  Terrence C. Stewart\,$^{1}$,
  Eric Hunsberger\,$^1$,
  Andrew Mundy\,$^2$ and
  Chris Eliasmith\,$^1$}
\def\Address{$^{1}$Centre for Theoretical Neuroscience,
  University of Waterloo,
  Waterloo, ON, Canada\\
  $^{2}$ Advanced Processors Technologies Research Group,
  School of Computer Science,
  University of Manchester,
  Manchester, UK}
\def\corrAuthor{Trevor Bekolay}
\def\corrAddress{Centre for Theoretical Neuroscience,
  David R. Cheriton School of Computer Science,
  University of Waterloo,
  200 University Avenue West,
  Waterloo, ON  N2L~3G1, Canada}
\def\corrEmail{tbekolay@uwaterloo.ca}

@@

Benchmarking involves collecting performance metrics
for two or more comparable systems.
Currently, the Nengo test suite collects
three types of performance metrics
across five backends.

@@

\paragraph{Compliance\textcolon}

\paragraph{Accuracy\textcolon}

\paragraph{Speed\textcolon}

@@

\paragraph{Reference\textcolon}

\paragraph{Simplified reference\textcolon}

\paragraph{OpenCL\textcolon}

\paragraph{SpiNNaker hardware\textcolon}

\paragraph{Neurogrid simulation\textcolon}

@@

@@

@@

%% Frontiers supports the policy of data sharing, and authors are advised
%% to make freely available any materials and information described in
%% their article, and any data relevant to the article (while not
%% compromising confidentiality in the context of human-subject research)
%% that may be reasonably requested by others for the purpose of academic
%% and non-commercial research. In regards to deposition of data and data
%% sharing through databases, Frontiers urges authors to comply with the
%% current best practices within their discipline.

@@

%Frontiers follows the recommendations by the International Committee
%of Medical Journal Editors
%(http://www.icmje.org/ethical_4conflicts.html) which require that all
%financial, commercial or other relationships that might be perceived
%by the academic community as representing a potential conflict of
%interest must be disclosed. If no such relationship exists, authors
%will be asked to declare that the research was conducted in the
%absence of any commercial or financial relationships that could be
%construed as a potential conflict of interest. When disclosing the
%potential conflict of interest, the authors need to address the
%following points: • Did you or your institution at any time receive
%payment or services from a third party for any aspect of the
%submitted work?  • Please declare financial relationships with
%entities that could be perceived to influence, or that give the
%appearance of potentially influencing, what you wrote in the
%submitted work.  • Please declare patents and copyrights, whether
%pending, issued, licensed and/or receiving royalties relevant to the
%work.  • Please state other relationships or activities that readers
%could perceive to have influenced, or that give the appearance of
%potentially influencing, what you wrote in the submitted work.

%% The authors declare that the research was conducted in the absence of
%% any commercial or financial relationships that could be construed as a
%% potential conflict of interest.

@@

%When determining authorship the following criteria should be observed:
%• Substantial contributions to the conception or design of the work; or the acquisition, analysis, or interpretation of data for the work; AND
%• Drafting the work or revising it critically for important intellectual content; AND
%• Final approval of the version to be published ; AND
%• Agreement to be accountable for all aspects of the work in ensuring that questions related to the accuracy or integrity of any part of the work are appropriately investigated and resolved.
%Contributors who meet fewer than all 4 of the above criteria for authorship should not be listed as authors, but they should be acknowledged. (http://www.icmje.org/roles_a.html)

%% The statement about the authors and contributors can be up to several
%% sentences long, describing the tasks of individual authors referred to
%% by their initials and should be included at the end of the manuscript
%% before the References section.

\section*{Acknowledgments}
%% Acknowledgments

@@

\section*{Supplemental Data}
%% Supplementary Material should be uploaded separately on submission, if
%% there are Supplementary Figures, please include the caption in the
%% same file as the figure. LaTeX Supplementary Material templates can be
%% found in the Frontiers LaTeX folder

@@

@@

%% \begin{table}[!t]
%% \textbf{\refstepcounter{table}\label{Tab:01} Table \arabic{table}.}{
%%   Maximum size of the Manuscript }

%% \processtable{ }
%% {\begin{tabular}{lllll}\toprule
%%  & Abstract max. legth (incl. spaces) & Figures or tables & Manuscript max. length & Final PDF length\\\midrule
%%  Clinical Case Study & & & &\\
%%  Clinical Trial & & & &\\
%%  Hypothesis and Theory & & & &\\
%%  Methods & 2000 characters  & 15 & 12000 words & 12 pages\\
%%  Original Research & & & &\\
%%  Review & & & &\\
%%  Technology Report & & & &\\
%%  Focused Review & 2000 characters & 5 & 5000 words & 5 pages\\
%%  CPC &  1250 characters& 6 & 2500 words & 4 pages\\
%%  Perspective & 1250 characters & 2 & 3000 words & 3 pages\\
%%  Mini Review & & & &\\
%%  Classification & 1250 characters & 10 & 2000 words & 12 pages\\
%%  Editorial & none & none & 1000 words & 1 page \\
%%  Book review & & & &\\
%%  Frontiers Commentary & none & 1 & 1000 words & 1 page\\
%%  General Commentary & & & &\\
%%  Field Grand Challenge & & & &\\
%%  Opinion & none & 1 & 2000 words & 2 pages\\
%%  Specialty Grand Challenge& & & &\\\botrule
%% \end{tabular}}{}
%% \end{table}

%% \begin{figure}[h!]
%% \begin{center}
%% \includegraphics[width=10cm]{logo1}
%% \end{center}
%% \textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}
%%        { Enter the caption for your figure here.  Repeat as necessary
%%          for each of your figures }
%% \end{figure}

>>>>

\graphicspath{{../figures/}}

\def\keyFont{\fontsize{8}{11}\helveticabold }
\def\firstAuthorLast{Bekolay {et~al.}}
\def\Authors{Trevor Bekolay\,$^{1,*}$,
  Terrence C. Stewart\,$^{1}$, and
  Chris Eliasmith\,$^1$}
\def\Address{$^{1}$Centre for Theoretical Neuroscience,
  University of Waterloo,
  Waterloo, ON, Canada}
\def\corrAuthor{Trevor Bekolay}
\def\corrAddress{Centre for Theoretical Neuroscience,
  David R. Cheriton School of Computer Science,
  University of Waterloo,
  200 University Avenue West,
  Waterloo, ON  N2L~3G1, Canada}
\def\corrEmail{tbekolay@uwaterloo.ca}

@@

Benchmarking involves collecting performance metrics
for two or more comparable systems.
We collect three metrics in this study,
as a proof of concept that the Nengo test suite
can be used to collect meaningful benchmarks
for neural simulators and neuromorphic hardware.

@@

\paragraph{Compliance}
is the number of tests that a particular backend
passes successfully, relative to the reference
backend, which must pass all tests.
While this metric can give an indication
of how many features a particular backend supports,
it does not take into account how commonly used
a particular feature is;
for this reason, a backend that has relatively
low compliance can still be very useful
in many situations.

\paragraph{Accuracy}
measures how well a particular backend
implements a desired model.
The actual accuracy metric depends
on the desired model,
but in simple cases can be
the root mean squared error (RMSE)
between a desired signal
and the actual signal decoded
from an ensemble.

\paragraph{Speed}
measures the amount of time it takes
for a backend to run a model.
Primarily, we are concerned with
the amount of time taken
in \texttt{run} calls;
however, we also separately measure how long
each backend spends building each model
(i.e., the time spent in \texttt{\_\_init\_\_}),
as models must be built before being run.

\subsection{Test models}

Unlike the compliance metric,
accuracy and speed are only collected
for tests that define an an accuracy metric
(i.e., those that use the \texttt{analytics} fixture).
In this study, we focus on four such tests,
which we claim are all necessary components
of large-scale cognitive brain models,
though they are by no means a sufficient set
for all complex cognitive behaviors.

\paragraph{1. Communication channel chain.}
In this model,
five ensembles,
comprised of 500 leaky integrate-and-fire (LIF) neurons total,
are connected in series,
with a communication channel
(i.e., the identity function)
computed across each connection.
This model, therefore, attempts to faithfully
communicate an input signal to
the last ensemble in the chain.
The input signal used is a static vector.
This model tests how
robust each backend is to noise introduced
by representing a vector space
in the activity of spiking neurons.

\paragraph{2. Two-dimensional product.}
In this model,
the scalar product is computed
from a two-dimensional ensemble
comprised of 100 LIF neurons.
Since the product is nonlinear,
we use a space-filling curve
(the Hilbert curve \cite{???})
as the input signal to ensure that
we sample the entire two-dimensional space.
This model tests each backend's
ability to compute a nonlinear function,
albeit a low-dimensional one.
Despite the low dimensionality,
two-dimensional products are frequently used
in large-scale models;
it is the primary operation used to
bind semantic pointers, for example
\cite{???}.

\paragraph{3. Controlled oscillator.}
In this model,
a three-dimensional ensemble is recurrently connected
such that the first two dimensions continuously
traverse a limit cycle
(i.e., they implement a cyclic attractor),
and the third dimension controls the speed
and direction of the oscillation.
As input, we provide an initial stimulus
to start the oscillation,
and provide a control signal
to cause oscillation at
2Hz, 1Hz, 0Hz, -1Hz, and -2Hz.
In total, the model is comprised of 600 LIF neurons.
This model tests each backend's ability
to stably implement a dynamical system,
which is required for many cognitive functions
like working memory and motor control.

\paragraph{4. Basal Ganglia sequence.}
In this model, comprised of 4900 LIF neurons,
a basal ganglia is created (as described in \cite{???})
that runs through a fixed sequence of actions.
Stewart et al. \cite{???} have showed that
these simple action sequences should
switch at a rate of ???.
This model tests that each backend can
build this basal ganglia model,
which is commonly used for action selection,
and that actions can be selected at the
hypothesized rate.

@@

We collected the three benchmark metrics
on five Nengo backends.

\paragraph{1. Reference (\texttt{nengo}).}
The reference backend is designed to run quickly
on any general purpose computer
by using NumPy \cite{???}
for fast vectorized computations.
It is included with the \texttt{nengo} frontend
as \texttt{nengo.Simulator}.
When appropriate, other backends will be benchmarked
relative to the reference backend,
which has the most features but does not target
specialized hardware.

\paragraph{2. Distilled (\texttt{nengo\_distilled}).}
The distilled backend is intended as a teaching tool,
and as a template for building new backends.
It can also run on any general purpose computer,
and also uses NumPy for fast vectorized computations,
but does not aim to implement all of the features
implemented by the reference backend,
and omits some optimizations that obfuscate code.
Therefore, the distilled backend is easier to read
and suitable for learning about Nengo
or as a starting point for new backends.

\paragraph{3. OpenCL (\texttt{nengo\_ocl}).}
The OpenCL backend uses the Open Computing Language
(OpenCL; \cite{???})
to run Nengo models on many different computing devices,
including graphical processing units (GPUs),
and field-programmable gate arrays (FPGAs).
In contrast to the distilled backend,
it is designed to run Nengo models
as quickly as possible,
using fast general purpose computing devices
like GPUs and any optimizations available,
at the cost of code readability.

\paragraph{4. Neurogrid software (\texttt{nengo\_brainstorm}).}
The Neurogrid backend is a software implementation
of a potential neuromorphic chip being developed
by Boahen et al. \cite{???}.
It does not aim for speed;
instead, it attempts to emulate the hardware
in order to test its applicability
to large-scale neural models.
If the emulated hardware can perform well,
then it follows that actual hardware
will also perform well,
but will be much faster.

\paragraph{5. SpiNNaker hardware (\texttt{nengo\_spinnaker}).}
The SpiNNaker backend targets
the eponymous neuromorphic hardware
developed by Furber et al. \cite{???}.
In contrast to the Neurogrid backend,
this backend targets physical neuromorphic hardware,
and therefore is concerned
both with accuracy and speed.

\subsection{Hardware used for running benchmarks}

??? ask Xuan

SpiNN-5 board

@@

\subsection{Compliance}

Table~\ref{Tab:01} gives compliance results for
all backends except the SpiNNaker backend,
which is not yet set up to run the Nengo test suite.
Of the four currently running unit tests,
the OpenCL backend has the highest compliance
(other than the reference backend).
The distilled backend has lower compliance
than OpenCL, which follows from its stated goal
of being simple but not necessarily feature-rich.
The Neurogrid backend is the least compliant,
but it is still an early experimental test bed;
this is not necessarily indicative of the
final Neurogrid hardware backend being
less compliant than other backends.

\subsection{Accuracy}

Figure~\ref{fig:01} gives accuracy results
for each backend on the chained communication channel model.
The boxplot shows that all five backends can
implement this model accurately,
despite five sources of potential noise.

Figure~\ref{fig:02} gives accuracy results
for the two-dimensional product model.
In this case, while all of the backends
perform well, the SpiNNaker backend
performs significantly worse than the
other four backends.
Its performance, however, is still well within
acceptable ranges for the two-dimensional product model.

Figure~\ref{fig:03} gives accuracy results
for the controlled oscillator model.
The reference and OpenCL backends perform well here,
as do the distilled and Neurogrid backends.
The SpiNNaker backend performs well in most cases,
but some instances of the model
are significantly less accurate
than the majority of instances.

Figure~\ref{fig:04} gives accuracy results
for the basal ganglia sequence model.
Again, all of the backends perform similarly,
except for the SpiNNaker backend.
All other backends have a transition time
around 43 milliseconds, which matches
the results from Stewart et al. 2010 \cite{???}
for simple actions.
The SpiNNaker backend has a transition time
around 55 milliseconds, which is slower
than the time reported by Stewart et al.
for simple actions, but faster than the
time reported for complex actions.

\subsection{Speed}

Figures~\ref{fig:05} and \ref{fig:06} show
the build and run speeds for all backends, respectively.
While build time is not critical to optimize,
it is worth noting that the reference backend
has consistently fast builds.
The OpenCL backend is also usually fast,
but can be slow for certain types of networks
(e.g., the chained communication channel model).
The distilled and Neurogrid backends---which
have a very similar build process---become
slow when dealing with moderately sized models
(e.g., the basal ganglia sequence model,
which contains 4900 neurons).
The SpiNNaker backend incurs an unavoidable cost
in setting up the hardware,
though it builds the largest model faster than
the distilled and Neurogrid backends.

The run time results are of the utmost importance
when evaluating backends for different applications,
especially those that require real-time interaction,
such as robotics. In general, all backends
perform adequately for the three smaller models.
The OpenCL backend is significantly slower for small
models, like the product and controlled oscillator models.
SpiNNaker, on the other hand, is slower than other backends
for these models, but only because its goal
is to run in real-time even if it is possible to run faster.
(??? Andrew is this true?)
The most important result is for the largest model,
the basal ganglia sequence.
In this model, the distilled and Neurogrid backends
performed very poorly
(though neither has speed as a primary goal).
The reference backend also did not perform well,
operating at nearly eight times slower than real-time.
The OpenCL backend performed well, operating at around
four times slower than real-time.
The best performance was seen by the SpiNNaker backend,
however, which operated at around 1.5 times real-time
on this moderately-sized model.

Despite the fast run times for
the OpenCL and SpiNNaker backends,
we also investigated the possibility
that probes were slowing these backend down,
since probes require moving data from
their native platform (GPU or SpiNNaker board)
to the CPU. However, we found that
removing probes from these models
made almost no impact on their run time
(see Figure~\ref{Fig:07}).

@@

The most important finding from these benchmarks
is that specialized hardware (including GPUs)
can result in significantly faster run times
for certain models,
with little to no cost in accuracy.
The conditions under which the OpenCL backend
performs slower than the reference backend
(e.g., the product and controlled oscillator models)
bear closer inspection to determine the source
of the slowdown.
Similarly, we should investigate
the model instances in which the SpiNNaker backend
is less accurate (e.g., in the controlled oscillator model)
or different from other backends
(e.g., in the basal ganglia sequence model).

Above and beyond the benchmarking results,
however, we assert that the primary contribution
of this study is to provide evidence that Nengo
has a tested, stable, productive frontend
that can target multiple backends,
and therefore provides a platform
for benchmarking neuromorphic backends,
and other neural simulators.
While this is similar to the goal
of similar projects, such a PyNN \cite{???},
Nengo is unique in focusing on large-scale simulations,
rather than attempting to support
detailed single-neuron models
(though this capability is still possible in Nengo;
see \cite{???}).

While we cannot claim that Nengo
solves all of complications that arise
in benchmarking,
we believe that it improves upon
the three major issues identified in the introduction.
First, benchmarking neuromorphic hardware
with Nengo is less biased than
hardware-specific benchmarks,
because the purpose of Nengo is
to make functionally interesting large-scale models.
These models are typically build and tested
in the reference backend,
leaving little opportunity to introduce
hardware-specific optimizations
that can bias benchmarks.

Second, there is less wasted effort
in using Nengo for benchmarking,
as the only requirement for using
Nengo is to develop a backend.
Building a Nengo backend for
a piece of neuromorphic hardware
gives access to all of the models
already developed for it,
which is reason enough to develop a backend.
The ability to collect functionally relevant benchmarks
comes ``for free'' once the backend exists.

Third, these benchmarks are more likely to
remain up-to-date because backend developers
themselves do not have to implement or update
these benchmarks;
they will be implemented to test Nengo's
capabilities as a neural simulator,
as an extension of the existing test suite.
Once implemented, backends that implement
the features that are used in a particular benchmark
should not have to modify their own code
to run those benchmarks on their own hardware.

???concluding remarks

@@

???put results.zip on figshare

???link to Github

@@

???probably something about ABR here?

@@

???todo

\section*{Acknowledgments}

???todo
(Andrew Mundy for SpiNNaker help, feedback on first draft,
Xuan for GPU server, Eric for help in getting OCL benchmarks running,
James for implementing most of OCL, Jan for product benchmark)

@@

@@

\begin{table}[!ht]
\textbf{\refstepcounter{table}\label{Tab:01} Table \arabic{table}.}{
  Compliance of four backends.}

\processtable{ }
{\begin{tabular}{lcc}\toprule
 Backend & Number of tests passed & Percentage of tests passed\\\midrule
 Reference (\texttt{nengo}) & 212 & 100\% \\
 Distilled (\texttt{nengo\_distilled}) & 129 & 60.8\% \\
 OpenCL (\texttt{nengo\_ocl}) & 187 & 88.2\% \\
 Neurogrid (\texttt{nengo\_brainstorm}) & 107 & 50.4\% \\\botrule
\end{tabular}}{}
\end{table}

\begin{figure}[!ht]
\begin{center}
  \includegraphics[width=85mm]{fig1}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:01} Figure \arabic{figure}.}
       {Chained communication channel model. (Top) One example instance
         of the model run on the reference backend; the decoded output
         of the five chained two-dimensional ensembles during the first
         12 milliseconds of the simulation are shown.
         Each ensemble reaches approximately the correct value of
         $[-0.6, 0.6]$ (indicated with the solid black line)
         but with slight delays depending on their position in the chain.
         (Bottom) The aggregated accuracy results for each backend on the
         chained communication channel model. The observed metric is the root
         mean-squared error (RMSE) between the decoded output of the last
         ensemble in the chain and the correct value (which is randomly
         selected on each trial), during the last 100 milliseconds of the
         500 millisecond simulation. The box shows the median and
         inter-quartile range, while the whiskers extend to 1.5 times the
         inter-quartile range; outliers are not shown.}
\end{figure}

\begin{figure}[!ht]
\begin{center}
  \includegraphics[width=85mm]{fig2}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:02} Figure \arabic{figure}.}
       {Two-dimensional product model. (Top) One example instance of the model
         run on the reference backend; the decoded input represented
         by the two-dimensional ensemble is shown. The input signal is
         designed to fully explore the two-dimensional vector space
         in the range $[-1, 1]$.
         (Middle) The decoded product represented by an ensemble downstream
         of the ensemble representing the input. It closely matches the
         correct product (indicated with the solid black line).
         (Bottom) The aggregated accuracy results for each backend on the
         two-dimensional product model. The measured metric is the root
         mean-squared error (RMSE) between the decoded output of the ensemble
         representing the product and the correct product, during the whole
         range of inputs provided over 5 simulated seconds (after a 500
         millisecond wait period to allow the input ensemble to reach the
         correct initial state). The box shows the median and
         inter-quartile range, while the whiskers extend to 1.5 times the
         inter-quartile range; outliers are not shown.}
\end{figure}

\begin{figure}[!ht]
\begin{center}
  \includegraphics[width=180mm]{fig3}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:03} Figure \arabic{figure}.}
       {Controlled two-dimensional oscillator model. (Left) One example
         instance of the model run on the reference backend; the decoded
         value represented by one of the two dimensions in the oscillatory
         ensemble is shown. The control signal dictating the frequency of
         oscillation changes every two seconds; in this plot each two second
         period is overlaid. Each signal oscillates at approximately the
         correct frequency (see middle for legend).
         (Middle) The power spectrum of each two second period.
         High power is seen only at the frequency we attempt to induce,
         for each of the five test frequencies.
         (Right) The aggregated accuracy results for each backend on the
         two-dimensional product model. The measured metric is the similarity
         between the power spectrum (calculated using the Fast Fourier
         Transform or FFT) of the decoded output of one of the two oscillatory
         dimensions in the ensemble, and the ideal power spectrum,
         computed on a pure sine wave oscillating at the induced frequency.
         Similarity is computed using a normalized dot product across
         all frequencies. The box shows the median and
         inter-quartile range, while the whiskers extend to 1.5 times the
         inter-quartile range; outliers are not shown.}
\end{figure}

\begin{figure}[!ht]
\begin{center}
  \includegraphics[width=85mm]{fig4}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:04} Figure \arabic{figure}.}
       {Basal Ganglia sequence model. (Top) One example instance
         of the model run on the reference backend; the similarity between
         the current representation in working memory and the six possible
         actions is shown (note that in the full benchmark, ten items are
         used). The model quickly progresses from one action to the next,
         cycling back at the end of the six items. The point at which the
         model switches from selecting one action to another action
         (the transition times) are indicated with dashed grey lines.
         (Bottom) The aggregated accuracy results for each backend on the
         basal ganglia sequence model. The observed metric is the mean
         time to transition from one action to the next. The model is run
         for four second, allowing for approximately 90 transitions, depending
         on the transition time. The box shows the median and
         inter-quartile range, while the whiskers extend to 1.5 times the
         inter-quartile range; outliers are not shown.}
\end{figure}

\begin{figure}[!ht]
\begin{center}
  \includegraphics[width=85mm]{fig5}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:05} Figure \arabic{figure}.}
       {Build time for each model on each backend. Each bar represents
         the mean build time across 150 instances of each model;
         error bars are 95\% confidence intervals.}
\end{figure}

\begin{figure}[!ht]
\begin{center}
  \includegraphics[width=85mm]{fig6}
\end{center}
\textbf{\refstepcounter{figure}\label{fig:06} Figure \arabic{figure}.}
       {Run time for each model on each backend. Each bar represents
         the mean run time across 150 instances of each model;
         error bars are 95\% confidence intervals.
         Run time is measured relative to real time;
         e.g., a value of two indicates that the model ran twice as slow
         as real time.}
\end{figure}

@@
